Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2023-03-10 analysis/manuscript.qmd
Make website project and manuscript subproject targets
Use {targets} to render fancy and manuscripty versions of the paper through Quarto profiles!
THIS IS SO COOL!

Once https://github.com/quarto-dev/quarto-r/issues/95 and/or https://github.com/ropensci/tarchetypes/issues/139 get fixed this will be a ton easier too

<p class="text-center"><a class="btn btn-success" href="/manuscript/output/manuscript.html" target="manuscript">{{< fa file-lines >}} &ensp;HTML</a> <a class="btn btn-success" href="/manuscript/output/manuscript.html" target="_blank">{{< fa arrow-up-right-from-square >}} &ensp;HTML (new window)</a> <a class="btn btn-success" href="/manuscript/output/manuscript.pdf" target="manuscript">{{< fa file-pdf >}} &ensp;PDF</a> <a class="btn btn-success" target="_blank" href="/manuscript/output/manuscript.pdf">{{< fa file-arrow-down >}} &ensp;PDF (download)</a></p>

<iframe src="/manuscript/output/manuscript.html" frameborder="0" style="overflow:hidden;height:90%;width:100%;border:2px solid #ccc;" height="90%" width="100%" name="manuscript"></iframe>

>>>>

```{r include=FALSE}
# Add the manuscript target as a dependency to the overall website
targets::tar_load(c(manuscript_nice, manuscript_manuscripty))
```

```{=html}
<p class="text-center">
  <a class="btn btn-success btn-sm mb-1" href="/manuscript/output/manuscript.html" target="manuscript">
  <i class="fa-solid fa-file-lines"></i> &ensp;HTML
  </a>

  <a class="btn btn-success btn-sm mb-1" href="/manuscript/output/manuscript.html" target="_blank">
  <i class="fa-solid fa-arrow-up-right-from-square"></i>&ensp;HTML <small>(new window)</small>
  </a>

  <a class="btn btn-primary btn-sm mb-1" href="/manuscript/output/manuscript.pdf" target="manuscript">
  <i class="fa-solid fa-file-pdf"></i>&ensp;PDF
  </a>

  <a class="btn btn-primary btn-sm mb-1" target="_blank" href="/manuscript/output/manuscript.pdf">
  <i class="fa-solid fa-file-arrow-down"></i>&ensp;PDF <small>(download)</small>
  </a>

  <a class="btn btn-info btn-sm mb-1" href="/manuscript/output-ms/manuscript-ms.pdf" target="manuscript">
  <i class="fa-solid fa-file-pdf"></i>&ensp;Manuscripty PDF
  </a>

  <a class="btn btn-info btn-sm mb-1" target="_blank" href="/manuscript/output-ms/manuscript-ms.pdf">
  <i class="fa-solid fa-file-arrow-down"></i>&ensp;Manuscripty PDF <small>(download)</small>
  </a>
</p>

<iframe src="/manuscript/output/manuscript.html" frameborder="0" style="overflow:hidden;height:90%;width:100%;border:2px solid #ccc;" height="90%" width="100%" name="manuscript"></iframe>
```

### 2023-03-10 manuscript/manuscript.qmd
Use {targets} to render fancy and manuscripty versions of the paper through Quarto profiles!
THIS IS SO COOL!

Once https://github.com/quarto-dev/quarto-r/issues/95 and/or https://github.com/ropensci/tarchetypes/issues/139 get fixed this will be a ton easier too

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
subtitle: "Research note"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
abstract: |
  In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. Did states actually end up complying with their pandemic-era derogations? Did derogating and non-derogating states implement emergency measures differently, and if so, how? Using data from the Varieties of Democracy PanDem dataset and the Oxford Covid-19 Government Response Tracker, we find that states which derogated from their international human rights obligations imposed emergency measures that were temporary and non-discriminatory. On the other hand, states that did not derogate were more likely impose discriminatory measures, enact emergency measures without time limits and violate non-derogable rights. Our results support the role that flexibility mechanisms such as derogations play in international law.
date: March 8, 2023
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: bib.bib
link-citations: true
---

```{r setup, include=FALSE}
fmt_out <- knitr::pandoc_to()

@@

```{r libraries-data}
library(tidyverse)
library(scales)
library(lubridate)
library(tidybayes)
library(marginaleffects)
library(patchwork)
library(colorspace)
library(modelsummary)
library(kableExtra)
library(targets)

@@

@@

Using data from the Varieties of Democracy PanDem dataset, the Oxford Covid-19 Government Response Tracker, and United Nations treaty derogation data, we first assess the variation in human rights responses among governments during the pandemic. We then analyze the effect of derogating (or not) on human rights practices during the pandemic and assess the role of civil society in mitigating the effects of rights-violating policies. We find that states that derogated and specified time limits to the emergency provisions generally stuck to the time limits and adhered to standards required during derogation by avoiding discriminatory policies, the abusive enforcement of these policies, and not violating non-derogable rights. On the other hand, states that did not derogate from their international human rights obligations and still imposed emergency measures were more likely impose discriminatory policies, violate non-derogable rights (the right to life, freedom from torture, and freedom from slavery), and impose emergency measures without time limits.

@@

Recognizing that public health exigencies would require states to curb civil and political rights, the UN Human Rights Office of the High Commissioner (OHCHR) published guidance stating that any emergency measures must meet specific criteria of legality, necessity, proportionality, and non-discrimination. However,not every country formally derogated—a number of countries implemented emergency measures without formally derogating from ICCPR. Over 100 countries issued emergency declarations while fewer than 30 submitted COVID-19 related derogations [@icnl-covid]. These COVID-19 emergency declarations may have longer term impacts beyond temporary human rights violations [@BrownBrechenmacherCarothers:2020]. Below, we first theorize how derogations can protect against rights violations and the excesses of emergency measures, followed by the role of civil society in mitigating these violations.

@@

We expect that states that did not notify the ICCPR of impending rights suspensions and submit derogations were less likely to take the the criteria for derogations—necessity, proportionality, non-discrimination and time delimited seriously, and therefore more likely to impose measures that violated these criteria. If states determined that emergency measures are justified without consideration of international law, we expect that there will be more violations.

@@

The first few months of the pandemic raised numerous concerns that the pandemic would contribute to violations of democratic standards and human rights in a number of countries and would reinforce the powers of repressive governments worldwide.

Derogations are an international legal means to communicate to both the UN and global community that certain rights will be suspended during times of crisis. This research note set out to examine state responses to their international human rights treaty obligations, particularly the ICCPR, during the COVID-19 pandemic and any subsequent impact on human rights violations.

@@

Looking back to our driving questions, what we find is that there is good news for compliance when we look to derogations. Derogators did not systematically use this legal "escape clause" opportunistically—derogators used the process correctly. Most derogations included time limitations, states tended to adhere to the time limits, and there was not an overall abuse of the derogation time to abuse, repress, and discriminate. Though not many states submitted COVID-19 related derogations, the ones that did, it seems, used derogations as intended by international law. Once the time of derogation ended, states readjusted to typical compliance once more. States, therefore complied with the intent of derogations and the purpose of the treaty. This is an important finding for international legal behavior and compliance studies. Even in a time of crisis, overall, states considered it important to follow through with the processes and expectations of international human rights law bureaucracy.

We suggest that future research look to differences in levels within states during crisis. There may be key differences in enforcement between local level bureaucrats and higher level elites. While higher level diplomats submit derogations and may be more cautious about international law, local level enforcement may not be as concerned about international legal aspects such as time limits and derogable vs. non-derogable rights. This distinction may explain some variation in policy and discrimination practices during the pandemic.

We suggest future research look to other times of crisis and derogation use to compare COVID-19 area derogation compliance with that in other crises. This note contributed to the study of derogations by building on the analysis between derogation submission and outcome. States did end up taking the restriction measures they signaled they would via the derogations.

@@

```{r tbl-results-full-prelim}
#| tbl-cap: Results from models showing relationship between derogations and emergency policies
#| include: true
#| echo: false
notes <- c(
  "Estimates are median posterior log odds from ordered logistic and binary logistic regression models;",
  r"(95\\% credible intervals (highest density posterior interval, or HDPI) in brackets.)",
  "Total $R^2$ considers the variance of both population and group effects;",
  "marginal $R^2$ only takes population effects into account."
)

modelsummary(models_tbl_prelim,
             estimate = "{estimate}",
             statistic = "[{conf.low}, {conf.high}]",
             coef_map = coef_map,
             gof_map = gof_map,
             output = "kableExtra",
             fmt =  list(estimate = 2, conf.low = 2, conf.high = 2),
             escape = FALSE) %>% 
  kable_styling(htmltable_class = "table-sm light-border",
                font_size = ifelse(fmt_out == "latex", 8, NA)) %>% 
  footnote(general = notes, footnote_as_chunk = FALSE, escape = FALSE) %>% 
  column_spec(1:6, width = "0.16\\\\textwidth")
```

@@

```{r tbl-results-full-h1}
#| tbl-cap: Complete results from models showing relationship between derogations and human rights (H~1~)
#| include: true
#| echo: false
modelsummary(models_tbl_h1,
             estimate = "{estimate}",
             statistic = "[{conf.low}, {conf.high}]",
             coef_map = coef_map,
             gof_map = gof_map,
             output = "kableExtra",
             fmt =  list(estimate = 2, conf.low = 2, conf.high = 2),
             escape = FALSE) %>% 
  kable_styling(htmltable_class = "table-sm light-border",
                font_size = ifelse(fmt_out == "latex", 8, NA)) %>% 
  footnote(general = notes, footnote_as_chunk = FALSE, escape = FALSE) %>% 
  column_spec(1:5, width = "0.18\\\\textwidth")
```

@@

```{r tbl-results-full-h2}
#| tbl-cap: Models showing relationship between derogations, civil society, and human rights (H~2~)
#| include: true
#| echo: false
modelsummary(models_tbl_h2,
             estimate = "{estimate}",
             statistic = "[{conf.low}, {conf.high}]",
             coef_map = coef_map,
             gof_map = gof_map,
             output = "kableExtra",
             fmt =  list(estimate = 2, conf.low = 2, conf.high = 2),
             escape = FALSE) %>% 
  kable_styling(htmltable_class = "table-sm light-border",
                font_size = ifelse(fmt_out == "latex", 8, NA)) %>% 
  footnote(general = notes, footnote_as_chunk = FALSE, escape = FALSE) %>% 
  column_spec(1:5, width = "0.18\\\\textwidth")
```

>>>>

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
short-title: "Pandemic Pass?"
subtitle: "Research note"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
abstract: |
  In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. Did states actually end up complying with their pandemic-era derogations? Did derogating and non-derogating states implement emergency measures differently, and if so, how? Using data from the Varieties of Democracy PanDem dataset and the Oxford Covid-19 Government Response Tracker, we find that states which derogated from their international human rights obligations imposed emergency measures that were temporary and non-discriminatory. On the other hand, states that did not derogate were more likely impose discriminatory measures, enact emergency measures without time limits and violate non-derogable rights. Our results support the role that flexibility mechanisms such as derogations play in international law.
date: March 8, 2023
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: bib.bib
link-citations: true
---

```{r setup, include=FALSE}
if (is.null(knitr::pandoc_to())) {
  fmt_out <- "interactive"
} else {
  fmt_out <- knitr::pandoc_to()
}

@@

```{r libraries-data}
library(tidyverse)
library(scales)
library(lubridate)
library(tidybayes)
library(marginaleffects)
library(patchwork)
library(colorspace)
library(modelsummary)
library(kableExtra)
library(gt)
library(targets)

@@

\begingroup\small\sffamily

::: {.callout-note}
## Scope of this paper

This paper is not a traditional research article with statistical tests of novel theories. Instead it is a research note aimed at examining unexplored relationships between derogations, human rights, and civil society. We look forward to feedback and ideas to make it a strong research note!

@@

Using data from the Varieties of Democracy PanDem dataset, the Oxford Covid-19 Government Response Tracker, and United Nations treaty derogation data, we first assess the variation in human rights responses among governments during the pandemic. We then analyze the effect of derogating (or not) on human rights practices during the pandemic and assess the role of civil society in mitigating the effects of rights-violating policies. We find that states that derogated and specified time limits to the emergency provisions generally stuck to the time limits and adhered to standards required during derogation by avoiding discriminatory policies and not violating non-derogable rights. On the other hand, states that did not derogate from their international human rights obligations and still imposed emergency measures were more likely impose discriminatory policies, violate non-derogable rights (the right to life, freedom from torture, and freedom from slavery), and impose emergency measures without time limits.

@@

Recognizing that public health exigencies would require states to curb civil and political rights, the UN Human Rights Office of the High Commissioner (OHCHR) published guidance stating that any emergency measures must meet specific criteria of legality, necessity, proportionality, and non-discrimination. However, not every country formally derogated—a number of countries implemented emergency measures without formally derogating from ICCPR. Over 100 countries issued emergency declarations while fewer than 30 submitted COVID-19 related derogations [@icnl-covid]. These COVID-19 emergency declarations may have longer term impacts beyond temporary human rights violations [@BrownBrechenmacherCarothers:2020]. Below, we first theorize how derogations can protect against rights violations and the excesses of emergency measures, followed by the role of civil society in mitigating these violations.

@@

We expect that states that did not notify the ICCPR of impending rights suspensions and submit derogations were less likely to take the criteria for derogations—necessity, proportionality, non-discrimination and time delimited seriously, and therefore more likely to impose measures that violated these criteria. If states determined that emergency measures are justified without consideration of international law, we expect that there will be more violations.

@@

The first few months of the pandemic raised numerous concerns that the pandemic would contribute to violations of democratic standards and human rights in a number of countries and would reinforce the powers of repressive governments worldwide. Derogations are an international legal means to communicate to both the UN and global community that certain rights will be suspended during times of crisis. This research note set out to examine state responses to their international human rights treaty obligations, particularly the ICCPR, during the COVID-19 pandemic and any subsequent impact on human rights violations.

@@

With the caveat that our results are by no means causal—we have not attempted to use any formal strategy to identify an unconfounded causal effect, and our findings are merely descriptive—we find that we find that there is good news for compliance and derogations. Derogators did not systematically use this legal "escape clause" opportunistically—derogators used the process correctly. Most derogations included time limitations, states tended to adhere to the time limits, and there was not an overall abuse of the derogation time to abuse, repress, and discriminate. Though not many states submitted COVID-19 related derogations, it seems that the ones that did used derogations as intended by international law. Once the time of derogation ended, states readjusted to typical compliance once more. States, therefore complied with the intent of derogations and the purpose of the treaty. This is an important finding for international legal behavior and compliance studies. Even in a time of crisis, overall, states considered it important to follow through with the processes and expectations of international human rights law bureaucracy.

We suggest that future research look to differences in levels within states during crisis. There may be key differences in enforcement between local level bureaucrats and higher level elites. While higher level diplomats submit derogations and may be more cautious about international law, local level enforcement may not be as concerned about international legal aspects such as time limits and derogable vs. non-derogable rights. This distinction may explain some variation in policy and discrimination practices during the pandemic. We also recommend that future research compare COVID-19 area derogation compliance with that in other crises. This note contributed to the study of derogations by building on the analysis between derogation submission and outcome. States did end up taking the restriction measures they signaled they would via the derogations. Finally, the relationship and interaction between civil society, human rights, and international law remains an important and ripe field for future research.

@@

```{r tbl-results-full-prelim}
#| tbl-cap: Results from models showing relationship between derogations and emergency policies
#| include: true
#| echo: false
notes <- c(
  "Estimates are median posterior log odds from ordered logistic and binary logistic regression models;",
  "95% credible intervals (highest density posterior interval, or HDPI) in brackets.",
  "Total \\(R^2\\) considers the variance of both population and group effects;",
  "marginal \\(R^2\\) only takes population effects into account."
)

if (fmt_out == "latex") {
  notes <- notes %>% 
    str_replace_all(r"[\\]", r"[\\\\]") %>% 
    str_replace_all("%", r"(\\\\%)")
} else {
  notes <- paste(notes, collapse = " ")
}

if (fmt_out == "latex") { 
  modelsummary(models_tbl_prelim,
               estimate = "{estimate}",
               statistic = "[{conf.low}, {conf.high}]",
               coef_map = coef_map,
               gof_map = gof_map,
               output = "kableExtra",
               fmt =  list(estimate = 2, conf.low = 2, conf.high = 2),
               escape = FALSE) %>% 
    kable_styling(htmltable_class = "table-sm light-border",
                  font_size = ifelse(fmt_out == "latex", 8, NA)) %>% 
    footnote(general = notes, footnote_as_chunk = FALSE, escape = FALSE) %>% 
    column_spec(1:6, width = "0.16\\\\textwidth")
} else {
  modelsummary(models_tbl_prelim,
               estimate = "{estimate}",
               statistic = "[{conf.low}, {conf.high}]",
               coef_map = coef_map,
               gof_map = gof_map,
               output = "gt",
               fmt =  list(estimate = 2, conf.low = 2, conf.high = 2)) %>% 
    tab_footnote(footnote = notes) %>% 
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_column_labels())
}
```

@@

```{r tbl-results-full-h1}
#| tbl-cap: Complete results from models showing relationship between derogations and human rights (H~1~)
#| include: true
#| echo: false
if (fmt_out == "latex") {
  modelsummary(models_tbl_h1,
               estimate = "{estimate}",
               statistic = "[{conf.low}, {conf.high}]",
               coef_map = coef_map,
               gof_map = gof_map,
               output = "kableExtra",
               fmt =  list(estimate = 2, conf.low = 2, conf.high = 2),
               escape = FALSE) %>% 
    kable_styling(htmltable_class = "table-sm light-border",
                  font_size = ifelse(fmt_out == "latex", 8, NA)) %>% 
    footnote(general = notes, footnote_as_chunk = FALSE, escape = FALSE) %>% 
    column_spec(1:5, width = "0.18\\\\textwidth")
} else {
  modelsummary(models_tbl_h1,
               estimate = "{estimate}",
               statistic = "[{conf.low}, {conf.high}]",
               coef_map = coef_map,
               gof_map = gof_map,
               output = "gt",
               fmt =  list(estimate = 2, conf.low = 2, conf.high = 2),
               escape = FALSE) %>% 
    tab_footnote(footnote = notes) %>% 
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_column_labels())
}
```

@@

```{r tbl-results-full-h2}
#| tbl-cap: Models showing relationship between derogations, civil society, and human rights (H~2~)
#| include: true
#| echo: false
if (fmt_out == "latex") {
  modelsummary(models_tbl_h2,
               estimate = "{estimate}",
               statistic = "[{conf.low}, {conf.high}]",
               coef_map = coef_map,
               gof_map = gof_map,
               output = "kableExtra",
               fmt =  list(estimate = 2, conf.low = 2, conf.high = 2),
               escape = FALSE) %>% 
    kable_styling(htmltable_class = "table-sm light-border",
                  font_size = ifelse(fmt_out == "latex", 8, NA)) %>% 
    footnote(general = notes, footnote_as_chunk = FALSE, escape = FALSE) %>% 
    column_spec(1:5, width = "0.18\\\\textwidth")
} else {
  modelsummary(models_tbl_h2,
               estimate = "{estimate}",
               statistic = "[{conf.low}, {conf.high}]",
               coef_map = coef_map,
               gof_map = gof_map,
               output = "gt",
               fmt =  list(estimate = 2, conf.low = 2, conf.high = 2),
               escape = FALSE) %>% 
    tab_footnote(footnote = notes) %>% 
    tab_style(style = cell_text(weight = "bold"),
              locations = cells_column_labels())
}
```
